Scaling rates of true polar wander in convecting planets and moons

### 2016-03-08 tpw_rate.tex
Small rephrasing
Address comments

First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
Second, there is the body-fixed (or geographic) frame, which corresponds roughly to a no-net-rotation or hotspot reference frame. 
Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
as described in Section~\ref{sec:rotational_deformation}.
The relative rotation between the inertial and geographic reference frame is given by the planetary rotation vector $\mitbf{\Omega}$.
Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
which we will refer to as the ``$\mathbf{E}$-frame.'' 
Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
rotating the $\mathbf{E}$-frame with respect to the geographic frame, which we will describe
by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This, as well, may be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, and $C_{ij}$ is the moment of inertia without that response.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formulism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
\label{eq:liouville_ode}
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
The rate of polar wander is the angular rate of change of the spin axis $\left| \dot{\mitbf{\omega}} \right| = \dot{\Theta}$ in the geographic reference frame.
This can be calculated by dotting Equation~\eqref{eq:liouville_ode} with itself:
\begin{equation}
\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
\label{eq:polar_wander_rate_tensor}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\dot{\Theta}^2 = 
&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
\label{eq:polar_wander_rate}
\end{aligned}
\end{equation}
This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
Two special cases of this equation are of note. First, if it is evaluated on the octahedral plane
(the plane with direction cosines all $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
\begin{equation}
\begin{aligned}
\dot{\Theta}^2
&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
\end{aligned}
\end{equation}
The second invariant of the stress deviator is commonly used in the theories of elasticity and plasticity
as a convenient scalar approximation of the shear stress in a system.
The second invariant of the moment of inertia deviator can be similarly viewed as a scalar estimate of the ``rotational stress.''

The second special case is if the rotation vector $\mitbf{\omega}$ lies in the plane of the $\lambda_1$ axis ($\phi = 0$).
Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
\begin{equation}
\vert \dot{\Theta} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
\label{eq:simple_milankovitch}
\end{equation}
The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

@@

In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name: it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, 
and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, but they do not affect the overall treatment of this scaling.

@@

We now seek to evaluate Equation~\eqref{eq:quasistatic_angular_momentum} in the perturbed, convecting state.
Hydrostatic balance (Equation~\eqref{eq:hydrostatic}) ensures that the integral over the reference shape $V_H$ vanishes when $\rho = \rho_0$.
Nonzero contributions arise from perturbations in the density field or from perturbations in the shape.
To make this dependence explicit, we split the shape into the reference volume $V_H$ and perturbations from it $\Delta V$,
and use Equation~\eqref{eq:eos} to define density perturbations.
Substituting this decomposition into Equation~\eqref{eq:quasistatic_angular_momentum} brings the integral into the form
\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV +  \\
&\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
\int_{\Delta V} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 0.  \\
\end{aligned}
\end{equation}
The first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma\equiv \alpha \Delta T$.
Removing these, we find
\begin{equation}
\begin{aligned}
\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 
-\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV.
\end{aligned}
\end{equation}
This equation may be identified with Equation~\eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table~\ref{tab:nondim}.

@@

Having drawn the connection between the angular momentum equation (Section~\ref{sec:rotational_dynamics}) and the linear momentum 
equation (Section~\ref{sec:internal}), we would like to apply insights from fluid dynamics and scaling of the 
linear momentum equation to the problem of TPW. We would therefore like to rewrite Equation~\eqref{eq:simple_milankovitch}
in terms of the nondimensional numbers identified in Section~\ref{sec:internal}.
Specifically, we will need estimates for $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$, and $\theta$.

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces($m$ \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m.
\end{equation}

The time constant $T_1$ is a viscous relaxation time of the planetary mantle. It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }.
\end{equation}
Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
\label{eq:scaled_rotation}
\end{equation}

@@

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
Therefore, an estimate of one allows an estimate of the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
\label{eq:T_series}
\end{equation}
Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
Orthogonality of the basis functions for the expansion results in the integral picking out degree-two 
spherical harmonics in the lateral dimensions, and only having power in the lowest few radial functions $R_n(r)$.
Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
\begin{equation}
\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

@@

There are two processes at work which control the evolution of $\theta$, its growth through perturbations in the convective moment and 
its decay via relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler})
The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
(here $\mitbf{\Psi}$ defines the rotation of the $\mathbf{E}$-frame in the geographic frame, see Figure~\ref{fig:reference_frames}).
We can then write the rate of change of $\mitbf{\omega}$ in the $\mathbf{E}$-frame (denoted by primes)
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
Equation~\eqref{eq:liouville_ode} then becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that $\phi=0$
(for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
The direction of $\Psi$ is given by colatitude $\alpha$ and longitude $\beta$, as shown in Figure~\ref{fig:reference_frames}.

We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ in terms of $\theta$, $\alpha$, and $\beta$:
\begin{equation}
\dot{\theta} = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta.
\label{eq:theta_competition}
\end{equation}
Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
and $\theta$ is completely determined by $\mitbf{\Psi}$
(where we have neglected the orientation factor $\sin \alpha \sin \beta$ for the purposes of scaling).
In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
at least occasionally, be larger than its decay.

An estimate for $\vert \mitbf{\Psi} \vert$ must concern the stability of the convective moment of inertia. 
Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, 
and $\theta$ will never be large.
However, if $\mathbf{E}$ is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
Given a random perturbation to that moment, we would like to give a bound on how big the perturbation to $\theta$ is. 
This may be done by application of a theorem due to \citet{davis1970rotation}.

@@

Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
Since it is a 2D model, the spin axis may be represented by a single angle.
When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
At $\sim$1 Gyr the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
an IITPW event. But there are several other events where the eigengap becomes small, and there are 
still large TPW events associated with them.

>>>>

First, there is the inertial, non-rotating frame corresponding to spatial coordinates that are fixed in time.
Second, there is the body-fixed (or geographic) frame.
By definition, the rotaiton of the body-fixed frame relative to the inertial frame is specified by $\mitbf{\Omega}$.
A terrestrial no-net-rotation or hotspot reference frame are common choices for the body-fixed frame.
Treatments of gravitational or rotational deformation of a planet are most naturally expressed in the body-fixed frame,
as described in Section~\ref{sec:rotational_deformation}.
Finally, there is the frame described by the principal axes of convective part of the moment of inertia $\mathbf{E}$,
which we will refer to as the ``$\mathbf{E}$-frame.'' 
Redistribution of mantle mass anomalies due to convection changes the principal axes of $\mathbf{E}$, 
rotating the $\mathbf{E}$-frame with respect to the geographic frame, which we will describe
by a rotation vector $\mitbf{\Psi}$ (see Section~\ref{sec:theta}).

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This, as well, may be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, and $C_{ij}$ is the moment of inertia without that response.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
\label{eq:liouville_ode}
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
This correspondence permits useful insights (see below).
Let $\dot{\Theta} = \vert \dot{\mitbf{\omega}} \vert$ denote the rate of polar wander.
Evaluating the scalar product $\dot{\Theta}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
\begin{equation}
\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
\label{eq:polar_wander_rate_tensor}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\dot{\Theta}^2 = 
&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
\label{eq:polar_wander_rate}
\end{aligned}
\end{equation}
This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.
Two special cases of this equation are of note. First, if $\dot{\Theta}^2$ is evaluated on the octahedral plane
(the plane with direction cosines all $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
\begin{equation}
\begin{aligned}
\dot{\Theta}^2
&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
\end{aligned}
\end{equation}
The second invariant of the stress deviator is commonly used in the theories of elasticity and plasticity
as a convenient scalar approximation of the shear stress in a system.
The second invariant of the moment of inertia deviator can be similarly viewed as a scalar estimate of the ``rotational stress.''

The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_2$ plane ($\phi = 0$).
Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
\begin{equation}
\vert \dot{\Theta} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
\label{eq:simple_milankovitch}
\end{equation}
The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

@@

In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name: it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, 
and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers that characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, but they do not affect the overall treatment of this scaling.

@@

We now seek to evaluate Equation~\eqref{eq:quasistatic_angular_momentum} in the perturbed, convecting state.
Hydrostatic balance (Equation~\eqref{eq:hydrostatic}) ensures that the integral over the reference shape $V_H$ vanishes when $\rho = \rho_0$.
Nonzero contributions arise from perturbations in the density field or from perturbations in the shape.
To make this dependence explicit, we split the shape into the reference volume $V_H$ and perturbations from it $\Delta V$,
and use Equation~\eqref{eq:eos} to define density perturbations.
Substituting this decomposition into Equation~\eqref{eq:quasistatic_angular_momentum} brings the integral into the form
\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV +  \\
&\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV + 
\int_{\Delta V} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 0.  \\
\end{aligned}
\end{equation}
As previously noted, the first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma\equiv \alpha \Delta T$.
Removing these, we find
\begin{equation}
\begin{aligned}
\int_{\Delta V} \rho_0 \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV = 
-\int_{V_H} \rho_0 \alpha (T-T_0) \mathbf{r} \times \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} \,dV.
\end{aligned}
\end{equation}
This equation may be identified with Equation~\eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table~\ref{tab:nondim}.

@@

Having drawn the connection between the angular momentum equation (Section~\ref{sec:rotational_dynamics}) and the linear momentum 
equation (Section~\ref{sec:internal}), we would like to apply the scaling of the 
linear momentum equation to the problem of TPW. We would therefore like to rewrite Equation~\eqref{eq:simple_milankovitch}
in terms of the nondimensional numbers identified in Section~\ref{sec:internal}.
Specifically, we will need estimates for $(C-A)$, $T_1$, $(\lambda_3-\lambda_1)$, and $\theta$.

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the ratio of rotational to gravitational forces(\citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m.
\end{equation}

The time constant $T_1$ is a viscous relaxation time of the planetary mantle. It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }.
\end{equation}
Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\Theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}.
\label{eq:scaled_rotation}
\end{equation}

@@

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
Therefore, an estimate of one allows an estimate of the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
\label{eq:T_series}
\end{equation}
Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
\begin{equation}
\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

@@

There are two processes at work which control the evolution of $\theta$, its growth through perturbations in the convective moment and 
its decay via relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
(here $\mitbf{\Psi}$ defines the rotation of the $\mathbf{E}$-frame in the geographic frame, see Figure~\ref{fig:reference_frames}).
We can then write the rate of change of $\mitbf{\omega}$ in the $\mathbf{E}$-frame (denoted by primes)
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
Equation~\eqref{eq:liouville_ode} then becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that $\phi=0$
(for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
The direction of $\Psi$ is given by colatitude $\alpha$ and longitude $\beta$, as shown in Figure~\ref{fig:reference_frames}.

We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ 
(i.e., the colatitudinal velocity in the $\mathbf{E}$ frame, not the TPW rate $\dot{\Theta}$)
in terms of $\theta$, $\alpha$, and $\beta$:
\begin{equation}
\dot{\theta} = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta.
\label{eq:theta_competition}
\end{equation}
Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
and $\theta$ is completely determined by $\mitbf{\Psi}$
(where we have neglected the orientation factor $\sin \alpha \sin \beta$ for the purposes of scaling).
In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
at least occasionally, be larger than its decay.

An estimate for $\vert \mitbf{\Psi} \vert$ must concern the stability of the convective moment of inertia. 
Convection is continuously redistributing mass throughout the mantle, which will perturb the convective moment of inertia.  
If the convective moment is stable to small perturbations, then $\vert \mitbf{\Psi} \vert$ should be small, 
and $\theta$ will never be large.
However, if $\mathbf{E}$ is not stable to perturbations, then $\theta$ could be up to $90^\circ$. 
Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of the perturbation to $\theta$. 
This may be done by application of a theorem due to \citet{davis1970rotation}.

@@

Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
Since it is a 2D model, the spin axis may be represented by a single angle.
When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
At $\sim$1 Gyr the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
an IITPW event. But there are several other events where the eigengap becomes small, and there are 
still large TPW events associated with them.
For example, at $\sim$0.3 Gyr the eigengap dips, with an associated large TPW event, even though
there is technically no interchange of the axes.
